Elo World, a framework for \\ benchmarking weak chess \\ engines

### 2019-03-13 chess/papers/weak.tex
Writing on the shuttle and waiting for the gate, describing some more players

\title[Elo World]{Elo World, a framework for \\ benchmarking weak chess \\ engines}
% \authorinfo{Dr. Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
\author{Dr. Tom Murphy VII Ph.D.}\email{tom7@tom7.org}
\thanks{Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
Appears in SIGBOVIK 19119 with the
%
title inflation
%
of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1600 rating points}

@@

\subsection{The players}

@@

\describeplayer{first\_move} Make the lexicographically first legal
move. The moves are ordered as $\langle$ \verb+src_row+,
\verb+src_col+, \verb+dst_row+, \verb+dst_col+,
\verb+promote_to+ $\rangle$ for white (rank $1$ is row $0$) and the rows
are reversed for black to make the strategy symmetric. Tends to
produce draws (by repetition), because knights and rooks can often
move back and forth on the first few files. \deterministic

@@

\describeplayer{worstfish}
\describeplayer{generous}
\describeplayer{safe}
\describeplayer{popular}
\describeplayer{dangerous}
\describeplayer{rare}
\describeplayer{no\_i\_insist}
\describeplayer{sym\_mirror\_x}
\describeplayer{huddle}
\describeplayer{survivalist}
\describeplayer{sym\_180}
\describeplayer{sym\_mirror\_y}
\describeplayer{suicide\_king}
\describeplayer{reverse\_starting}
\describeplayer{fatalist}
\describeplayer{blind\_yolo}
\describeplayer{cccp}
\describeplayer{equalizer}
\describeplayer{blind\_kings}
\describeplayer{swarm}
\describeplayer{min\_oppt\_moves}
\describeplayer{chessmaster.nes\_lv1}
\describeplayer{stockfish1m\_r32768}
\describeplayer{chessmaster.nes\_lv2}
\describeplayer{stockfish1m\_r16384}
\describeplayer{stockfish0}
\describeplayer{stockfish5}
\describeplayer{stockfish1m\_r8192}
\describeplayer{topple10k}
\describeplayer{stockfish15}
\describeplayer{stockfish10}
\describeplayer{stockfish1m\_r4096}
\describeplayer{stockfish20}
\describeplayer{topple1m}
\describeplayer{stockfish1m\_r2048}
\describeplayer{stockfish1m\_r1024}
\describeplayer{stockfish1m\_r512}
\describeplayer{stockfish1m\_r256}
\describeplayer{stockfish1m\_r64}
\describeplayer{stockfish1m}
\describeplayer{stockfish1m\_r128}

@@

% Dramatic failure of the "suicide king" strategy against
% CCCP.
1. d4 g5 2. Bxg5 Nc6 3. Bxe7 Kxe7 4. d5 Kd6
5. dxc6+ Kc5 6. Qd6+ Kc4 7. e4++

@@

>>>>

\title[Elo World]{Elo World, a framework for \\ benchmarking weak chess \\ engines}
% \authorinfo{Dr.~Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
\author{Dr.~Tom Murphy VII Ph.D.}\email{tom7@tom7.org}
\thanks{Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
Appears in SIGBOVIK 19119 with the
%
title inflation
%
of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1600 rating points}

@@

\subsection{Simple players}

@@

\describeplayer{first\_move} Make the lexicographically first legal
move. The moves are ordered as $\langle\!\!$ \verb+src_row+,
\verb+src_col+, \verb+dst_row+, \verb+dst_col+,
\verb+promote_to+ $\!\!\rangle$ for white (rank $1$ is row $0$) and the rows
are reversed for black to make the strategy symmetric. Tends to
produce draws (by repetition), because knights and rooks can often
move back and forth on the first few files. \deterministic

@@

\describeplayer{huddle} As white, make moves that minimize the total
distance between white pieces and the white king. Distance is the
Chebyshev distance, which is the number of moves a king takes to move
between squares. This forms a defensive wall around the king
(Figure~\ref{fig:huddle}).

\begin{figure}[ht]
\chessboard[setfen=2r4r/5n2/1p4p1/nP5p/1Ppk3P/2NPRNP1/1bRPPP2/2BQKB2 b - - 80 158,showmover=false]
\caption{\player{huddle} (white) checkmates \player{pacifist} on move
  158 with substantial help. Note that white's distal pawns have
  advanced; these are actually the same distance from the King as they
  are in their home row, since the distance metric includes diagonal
  moves. } \label{fig:huddle}
\end{figure}

\describeplayer{swarm} Like \player{huddle}, but instead move pieces
such that they are close to opponent's king. This is essentially an
all-out attack with no planning, and manages to be one of the
better-performing ``simple'' strategies. From the bloodbaths it
creates, it even manages a few victories against strong opponents
(Figure~\ref{fig:swarm}).

% XXX illustrate the moves here
\begin{figure}[ht]
\chessboard[setfen=rnbqkb1r/pp1ppp1p/7p/1Bp5/3PnP2/8/PPP3PP/RN1QK1NR b KQkq - 1 5,showmover=false]
\caption{\player{swarm} (white) vs.~\player{stockfish1m\_r4096}
  with black to move after 1.~d4 Nf6 2.~Bh6 gxh6 3.~f4 c5 4.~e4 Nxe4
  5.~Bb5. Black blunders 5\ldots f6??, which must have been a random
  move as \player{swarm} immediately wins with 6.~Qh5++.} \label{fig:swarm}
\end{figure}

\describeplayer{generous} Move so as to maximize the number of
opponent moves that capture our pieces, weighting by the value of the
offered piece ($\pawn = 1$, $\bishop = \knight = 3$, $\rook = 5$,
$\queen = 9$). A piece that can be catpured multiple ways is counted
multiple times.

\describeplayer{no\_i\_insist} Like \player{generous}, but be overwhelmingly
polite by trying to {\em force} the opponent to accept the gift of material.
There are three tiers: Avoid at all costs mating the opponent (and moreso
checkmating). Stalemate is polite, but it is more canonical for two
polite players to form a draw by repetition, from continually offering up
material to one another and declining it. Next, avoid situations where
the opponent can refuse our gift; among these, prefer the move where the
opponent must capture the highest value piece. Finally, prefer moves
where the expected value of the offered material (i.e.~against random play)
is highest. (This means that if there are three moves, and one captures a
rook but the others capture nothing, the value is $5/3$.) This strategy
almost never wins, but is not among the worst players, since it often forces
a draw by exchanging all its pieces.

\describeplayer{reverse\_starting} This player thinks that the board is
upside-down, and as white, tries to put its pieces where black pieces
start. This often causes some conflict since the black pieces are already
there, but can also end peacefully (Figure~\ref{fig:reversestarting}).

\begin{figure}[ht]
\chessboard[setfen=1NB1K2R/8/8/8/8/8/6p1/rnbqkbn1 b - - 19 64,showmover=false]
\caption{\player{reverse\_starting} draws against \player{reverse\_starting}
by repetition, both players having happily moved their surviving pieces into the
correct spots.} \label{fig:reversestarting}
\end{figure}

% TODO: Cyrillic tag or toast at the end
\describeplayer{cccp} Prioritize moves that Checkmate, Check, Capture
or Push, in that order. Push means to move pieces as deep as possible
into enemy territory (without any regard for their safety). Ties are
broken deterministically by the source/destination squares, so this
one is technically asymmetric. \deterministic \asymmetric

\describeplayer{suicide\_king} Take a random move that minimizes the
distance between the two kings. Putting one's king out in the open is
very unprincipled (Figure~\ref{fig:cccpsuicide}), but it does produce
enough pressure to win against unambitious opponents.

\begin{figure}[ht]
  \chessboard[setfen=r1bq1bnr/pppp1p1p/2PQ4/8/2k6/8/PPP1PPPP/RN2KBNR b KQ - 0 7,showmover=false]
\caption{\player{suicide\_king} (black) dramatically failing against
  \player{cccp}'s slightly more principled play, after 1.~d4
  g5 2.~Bxg5 Nc6 3.~Bxe7 Kxe7 4.~d5 Kd6 5.~dxc6+ Kc5 6.~Qd6+ Kc4.
  White delivers a discovered mate with 7.~e4++.} \label{fig:cccpsuicide}
\end{figure}

% XXX include examples of symmetric players, as they are aesthetic.
\describeplayer{sym\_mirror\_y} As white, try to maximize symmetry
when flipping vertically. Zero penalty for opposing e.g.~a \King with
a \king, but a small penalty when the pieces are not the same type,
and a larger penalty when they are not opposite coors. The starting
position is already symmetric this way, so this usually has the effect
of copying the opponent's moves when possible. As ``copy your
opponent's moves'' is a commonly known (but underspecified) strategy
discovered by many children, this player is close to being canonical.
However, it admits a bit too much arbitrary choice in the penalties
assigned.

\describeplayer{sym\_mirror\_x} As \player{sym\_mirror\_y}, but
maximize symmetry when flipping horizontally. This does not make much
chess sense, but can produce aesthetic arrangements
(Figure~\ref{fig:XXX}).

% XXX show self play perhaps?
\describeplayer{sym\_180} As \player{sym\_mirror\_y}, but maximize
symmetry under $180\deg$ rotation of the board. An emergent priority is
to ``castle'' with the king and queen to ``fix'' them.

\describeplayer{min\_oppt\_moves} Take a move that minimizes the
number of resulting legal moves for the opponent, breaking ties
randomly. This is a beautifully simple approach that generalizes many
chess principles: Occupying space reduces the destination squares
available to the opponent; capturing their pieces reduces the number
of their pieces that they can move; pinning pieces or checking the
king further reduces the legal moves; and mating the opponent is the
best possible move.\footnote{However note that it does not distinguish
  checkmate and stalemate, despite these having very different
  results.} Among the players in the paper, this one is Pareto
efficient in terms of its simplicity and effectiveness. \canonical

% XXX cite blog comment?
\describeplayer{equalizer} Prefer to move a piece that has been
moved the fewest times, and then prefer moving to a square that
has been visited the fewest times. Castling counts as moving
both the king and rook, and visiting both destination squares.
This is the first strategy described that keeps meaningful state.
\canonical \stateful

\subsection{Fate-based players}

If we keep state, then we can track the location of each piece as it
moves around the board (allowing us to distinguish the two rooks, or
follow a pawn as it promotes). We can then use statistics on the
average fates of each piece over hundreds of millions of games to
guide the moves. These statistics give us, for each piece (e.g.~the c2
pawn) and square, how likely it is to end the game on that square, and
how likely it is to be alive or dead there when the game
ends~\cite{survival}.

\describeplayer{safe} This strategy moves pieces towards squares where
they are likely to end the game alive. For this strategy and several
others, simply moving to maximize this score (e.g.~its sum or product
over all pieces) is very boring, since the score is almost always
maximized in the starting position. So this strategy actually makes
each move randomly, weighted by the total score of the resulting
positions. The scores are normalized (with the lowest-scoring move
receiving weight 0.0 and the highest 1.0) and then sampled according
to these weights. Without normalization, the play is almost identical
to uniformly random, since the weights of the resulting positions tend
to be very similar (dominated by the many pieces that don't move). But
it's pretty arbitrary.

\describeplayer{popular} Like \player{safe}, but the score for a
piece/square pair is the probability that the piece ends on that
square, whether it lives or dies. This player likes to follow the
crowd!

\describeplayer{dangerous} The dual of \player{safe}; the score is
the probability that the piece dies on that square. Note that a
king is said to die if it is checkmated. This player likes to
live life on the edge!

\describeplayer{rare} The dual of \player{dangerous}; the score
is one minus the probability of ending the game on that square. This
player has a thirst for adventure!

\describeplayer{survivalist} Like the above, but the score is the
ratio of the survival and death probabilities on the square. In the
data set, every piece ends alive or dead in every square (except
for the bishops, which can only legally occupy squares of their
color) at least 1000 times, so each ratio is defined. Here, the
sums of ratios have plenty of variability, and the highest
ratios are not so often on the starting squares. So with this
strategy, we simply do a weighted sample from the moves according
to their (non-normalized) scores.

\describeplayer{fatalist} The dual of \player{fatalist}; the score is
the ratio of the death and survival probability on the square. This
player knows that even if you win, you just have to keep playing over
and over again, so you might as well get it over with!

\subsection{Engine-based players}

...

\describeplayer{stockfish0}
\describeplayer{stockfish5}
\describeplayer{stockfish15}
\describeplayer{stockfish10}
\describeplayer{stockfish20}

\describeplayer{worstfish} (compare ``losing chess'')
uses stockfish20; we could use 1m (but slow), but it
probably doesn't benefit from this depth of analysis,
since the engine assumes the opponent is playing well.
To do even worse, we would want to play moves that
trap the opponent into beating us, as well as ones that 
avoid repetition draws

\describeplayer{chessmaster.nes\_lv1}
\describeplayer{chessmaster.nes\_lv2} \deterministic
\describeplayer{stockfish1m\_r32768}
\describeplayer{stockfish1m\_r16384}
\describeplayer{stockfish1m\_r8192}
\describeplayer{topple1m}
\describeplayer{topple10k}
\describeplayer{stockfish1m}

\subsection{Blind players}

\describeplayer{blind\_yolo}
\describeplayer{blind\_kings}
% \describeplayer{blind\_spycheck} (described in paper, not yet run)

(I guess we can also segue by discussing handicaps)
(end with this so that we can segue into dilution methods)

\subsection{Diultion methods}
% Stockfish1m_32768 loses to reverse_starting in 3 moves!
1. e4 g5 2. Ba6 f5 3. Qh5

\describeplayer{stockfish1m\_r4096}
\describeplayer{stockfish1m\_r2048}
\describeplayer{stockfish1m\_r1024}
\describeplayer{stockfish1m\_r512}
\describeplayer{stockfish1m\_r256}
\describeplayer{stockfish1m\_r64}
\describeplayer{stockfish1m\_r128}

@@

@@

\section{Conclusion}

The author would like to thank the anonymous referees
of the Special Interest Group on
B
O
V
In
Chess journal for their careful reviews.
